Online {\IterComp}

### 2017-09-14 prof.tex
hmm
prof

In this section, we describe how the computation of the work metric can be performed during runtime by means of instrumenting the code.
%In particular, we adapt the optimal algorithm proposed originally for block-frequency profiling~\citep{nahapetian73,knuth73,ball94}.
Because we define work as a linear equation on the block-frequency counters, it is possible to embed its computation into the execution of the program.
A naive instrumentation would consist basically of having a global counter that starts with the interception value, $\varepsilon$, and each basic block increments its own cost into the global counter.
Although this instrumentation is easily implemented, it imposes a significant overhead on the instrumented program.
However, it is possible to insert fewer probes by carefully placing the probes in a way that is possible to reconstruct the complete profiling information~\citep{knuth73,ball94}.

We adapt the optimal block-frequency profiling in order to perform the work profiling efficiently.
The proposed work profiling differs from the optimal block-frequency profiling as the latter occurs in two stages:
\textit{(i.) Before execution:} The code is instrumented with counters for each probe.
\textit{(ii.) After execution:} The information from the recorded probes is propagated in the CFGs of the program.
In contrast, our work profiling has a single counter, and it only requires the instrumentation before execution, without any post-processing of the recorded profiling.

The instrumented code is assumed to be generated before optimising the code.
This assumption is based on three key points:
\textit{(i.)} it guarantees that the work metric is independent of optimisation, i.e., the same input is always mapped to the same amount of work, regardless of the optimisation;
\textit{(ii.)} it simplifies the code generator;
\textit{(iii.)} it leverages from the optimisations to further improve the instrumentation code.
Having a work metric that is independent of optimisation is the most important reason for which we instrument the code before optimisations.

The optimal placement of the probes works by first computing the maximum spanning
tree based on a weighing that assigns a value to each edge in the CFG.
These weights can be obtained either by empirical measurements or heuristic estimations,
and their goal is to avoid probing in frequently executed edges.
Once we have the maximum spanning tree, probes are placed on every edge not in the spanning tree.
Figure~\ref{fig:cfg-example} shows an example of a CFG with a maximum spanning tree.
%represented by the black edges, while the edges highlighted in red represent the placement of the probes.

\begin{figure}[t]
\centering {
  \includegraphics[scale=0.75]{figs/cfg-example.pdf}\\\vspace{1ex}
  \resizebox{0.45\textwidth}{!}{
  %\scalebox{0.8}{
     \begin{minipage}{0.5\textwidth}
     Instrumented value for each probe $P_i$:
     \begin{align*}
     \omega(P_0) &= w(B_0) + w(B_1) + w(B_4) + w(B_6) + w(B_7) + w(B_8) + w(B_9)\\
     \omega(P_1) &= w(B_2) - w(B_1) - w(B_4) - w(B_6) - w(B_7) - w(B_8)\\
     \omega(P_2) &= w(B_5) - w(B_6)\\
     \omega(P_3) &= w(B_3) - w(B_4) - w(B_6) - w(B_7)
     \end{align*}
     \end{minipage}
  }
}
  \caption{Example of a CFG with its minimum spanning tree in black and the
   basic blocks highlighted in red represent the instrumented basic blocks with
   the placement of the probes.}
  \label{fig:cfg-example}
\end{figure}

In contrast to the naive instrumentation where each basic block records only its
own amount of work, with the optimal profiling, the instrumented basic blocks need
to record an aggregated value of work that also takes other basic blocks into
account.
Some probes speculatively assume a path of execution,
while other probes correct when these assumptions are wrong
(see for example $w(P_0)$ and $w(P_1)$ in Figure~\ref{fig:cfg-example}).

We compute the aggregated value of work for each probe in two steps:
\textit{(i.)} we propagate information about the probes through the edges of the CFG,
as symbolic expressions, and then \textit{(ii.)} we use this edge flows
to compose the aggregated value of the probes.

@@

\begin{figure}[h]
  \centering
  \includegraphics[scale=0.75]{figs/cfg-relax-example.pdf}
  \caption{Example of a CFG containing a loop and its decomposition into DAGs.
           The DAGs are the subgraphs within the dashed boundaries.}
  \label{fig:cfg-relax-example}
\end{figure}

>>>>

In this section, we describe how to measure work online. Total work is just the sum of the costs of all executed basic blocks, so it is
possible to embed its computation into the execution of the program. A naive instrumentation would use a global counter initialized at the
interception value, $\varepsilon$. Each basic block would then have only to increment this counter with the total cost of its instructions
every time it is executed. Although easy to implement, the naive approach introduces a significant overhead.

Previous work on basic block profiling has proposed optimal ways for instrumenting the code with as little overhead as possible without
losing any profiling information~\citep{knuth73,ball94}. The underlying idea is to instrument only a subset of the basic blocks of a
function and use the function's CFG to calculate execution counts for the rest. We can use the same idea for our purposes. Work profiling
and basic block profiling are similar, both need to count how many times each basic block was executed. The only important difference is
that we do not have to store the execution count of each basic block, only the total work.

Graph theory shows that we can calculate precisely the frequencies of all edges between basic blocks if we choose a spanning tree of the
function CFG and instrument the edges not in the tree. To make the instrumentation probe placement optimal, we choose the maximum spanning
tree, with the edge weights representing edge frequency estimates. As many high frequency edges as possible will be in the maximum spanning
tree and they will not be instrumented.

\begin{figure}[t]
\centering {
  \includegraphics[scale=0.75]{figs/cfg-example.pdf}\\\vspace{1ex}
  \resizebox{0.45\textwidth}{!}{
  %\scalebox{0.8}{
     \begin{minipage}{0.5\textwidth}
     Instrumented value for each probe $P_i$:
     \begin{align*}
     \omega(P_0) &= w(B_0) + w(B_1) + w(B_4) + w(B_6) + w(B_7) + w(B_8) + w(B_9)\\
     \omega(P_1) &= w(B_2) - w(B_1) - w(B_4) - w(B_6) - w(B_7) - w(B_8)\\
     \omega(P_2) &= w(B_5) - w(B_6)\\
     \omega(P_3) &= w(B_3) - w(B_4) - w(B_6) - w(B_7)
     \end{align*}
     \end{minipage}
  }
}
  \caption{Example of a CFG with its maximum spanning tree in black. Basic blocks and edges highlighted in red are instrumented.
    Instrumenting them is enough for calculating the total work performed by the whole CFG.}
  \label{fig:cfg-example}
\end{figure}

In contrast to the naive instrumentation where each basic block records only its own amount of work, with the optimal profiling the work
counter increments needs to take other basic blocks into account. Figure~\ref{fig:cfg-example} shows an example CFG, where the instrumented
blocks are marked with red. In this example, executing block $B_3$ implies also executing $B_1$ and not executing $B_4$. We compute the
aggregated value of work for each probe in two steps: \textit{(i.)} we propagate information about the probes through the edges of the CFG,
as symbolic expressions, and then \textit{(ii.)} we use these edge flows to compose the aggregated value of the probes.

@@

\begin{figure}[t]
  \centering
  \includegraphics[scale=0.75]{figs/cfg-relax-example.pdf}
  \caption{Example of a CFG containing a loop and its decomposition into DAGs.
           The DAGs are the subgraphs within the dashed boundaries.}
  \label{fig:cfg-relax-example}
\end{figure}
